---
title: ""
subtitle: ""
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: ""
date: today
date-format: long
number-sections: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Elaineyi1/Children_Health"
toc: true
bibliography: references.bib
---
# Introduction

# Data
## Overview
## Methodology and Measurement
```{r}
#| include: false
library(arrow)
library(here)
library(haven)
library(modelsummary)
library(tidyr)
library(janitor)
library(dplyr)
library(ggplot2)
```

```{r}
#| include: false
data_cleaned <- read_parquet(here::here("data", "02-analysis_data", "children_health.parquet"))

train_data <- read_parquet(here::here("data", "02-analysis_data", "train_data.parquet"))
test_data <- read_parquet(here::here("data", "02-analysis_data", "test_data.parquet"))
```

## Data Visualizations and Analysis
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-age
#| fig-cap: Distribution of Child's Age and Whether the Child Needs Treatment for Emotional Development Behavior

data_cleaned |> ggplot(mapping = aes(x = age, fill = child_emotion_develop)) +
  geom_bar() +
  labs(x = "Age", y = "Number of observations", fill = "Whether a Child Needs Treatment \nfor Emotional Development Behaviour") +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-binary1
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 1
#| fig-width: 7
#| fig-height: 2
#| fig-cap: Core Based Statistical Area (CBSA) Status refers to the classification of a geographic area based on its relationship to an urban center, where CBSA is an area that has at least one urban center with a population of 10,000 or more. 
#| fig-subcap: ["Child's Sex and Emotion Developemnt", "CBSA and Child's Emotional Development", "Tenure of Living Places and Child's Emotional Development"]

data_cleaned |> ggplot(aes(x=sex,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "Sex", y = "Proportion") +
  guides(fill = guide_legend(position = "bottom"))

data_cleaned |> ggplot(aes(x=cbsa,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "Area Status", y = "Proportion") +
  guides(fill = guide_legend(position = "bottom"))

data_cleaned |> ggplot(aes(x=tenure,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "Tenure", y = "Proportion") +
  guides(fill = guide_legend(position = "bottom"))
```

```{r}
#| label: fig-binary2
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 1
#| fig-width: 7
#| fig-height: 2
#| fig-cap: Limited ability refers to a child's restriction or inability to do the things that most children of the same age can do.
#| fig-subcap: ["Child's Race and Emotion Developemnt", "Child's English Level and Emotion Developemnt", "Child's Limited Ability and Need for Emotional Development Treatment"]

data_cleaned |> ggplot(aes(x=race,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "Race", y = "Proportion") +
  guides(fill = guide_legend(position = "right"))

data_cleaned |> ggplot(aes(x=english,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "English Level", y = "Proportion") +
  guides(fill = guide_legend(position = "right"))

data_cleaned |> ggplot(aes(x=child_limited_ability,fill = factor(child_emotion_develop))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "azure3", "No" = "azure4"))+
  coord_flip()+
  labs(fill = "Whether a Child Needs \nTreatment for Emotional \nDevelopment Behaviour", x = "Limited Ability", y = "Proportion") +
  guides(fill = guide_legend(position = "right"))

```
# Model
## Model Set-up
$$
\begin{aligned}
y_i | \pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \beta_0 + \beta_1 \times \mbox{Age Group}_i + \beta_2 \times \mbox{Race}_i + \beta_3 \times \mbox{English}_i + \beta_4 \times \mbox{Child Limited Ability}_i \\
\beta_0 & \sim \mbox{Normal}(0, 2.5) \\
\beta_1 & \sim \mbox{Normal}(0, 2.5) \\
\beta_2 & \sim \mbox{Normal}(0, 2.5) \\
\beta_3 & \sim \mbox{Normal}(0, 2.5) \\
\beta_4 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$

where:

- $y_i$ is the dependent variable, representing whether a child needs treatment for emotional development behavior, where the outcome is binary (yes or no).
- $\beta_0$ is the intercept term, representing the value when all the predictor variables are zero. It has a prior distribution that is normal with a mean of 0 and a standard deviation of 2.5.
- $\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$ represent the effects of age group, race, English proficiency, and child limited ability, respectively, on the log-odds of a child needing treatment for emotional development behavior, with each having a prior distribution of Normal(0, 2.5).


## Model Justification and Limitations

```{r}
#| include: false
model <- readRDS(file = here::here("models/model.rds"))
```

# Results
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-model
#| tbl-cap: Estimating Whether a Child Needs Treatment for Emotional Development Behaviour Based on Age Group, Race, English Proficiency, and Limited Ability

modelsummary(
  list(
    "Treatment for Emotion Developement" = model
  )
)
```

# Discussions

# Appendix
## Surveys and Sampling
## Data Cleaning
## Posterior Predictive Checks

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}
#| label: fig-race_and_cbsa
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3
#| fig-cap: Core Based Statistical Area (CBSA) Status refers to the classification of a geographic area based on its relationship to an urban center, where CBSA is an area that has at least one urban center with a population of 10,000 or more. 
 
data_cleaned |> ggplot(aes(x=race,fill = factor(cbsa))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Located outside CBSA" = "azure3","Located within CBSA" = "azure4")) +
  coord_flip()+
  labs(fill = "Core Based Statistical \nArea Status", x = "Race", y = "Proportion") +
  guides(fill = guide_legend(position = "right")) +
  ggtitle("
Proportion of Core Based Statistical \nArea Status by Race")
```

```{r}
#| label: fig-hh_kids
#| echo: false
#| warning: false
#| message: false
data_transformed <- data_cleaned |>
  group_by(hhids) |>
  mutate(
    total_kids = n(),
    mental_health_count = sum(child_emotion_develop == "Yes", na.rm = TRUE),
    mental_health_proportion = mental_health_count / total_kids
  ) |>
  filter(!is.na(child_emotion_develop)) |>
  filter(mental_health_count > 0) 

ggplot(data_transformed, aes(x = mental_health_proportion)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    x = "Proportion of Children with Mental Health Issues",
    y = "Count of Households",
    title = "Proportion of Children with Mental Health Issues in \nHouseholds with Mental Health Issues"
  ) +
  scale_x_continuous(labels = scales::percent)
```

\newpage
# References